Synthesizing Verified Components for Cyber Assured Systems Engineering \thanks{{DISTRIBUTION STATEMENT A. Approved for public release.} }

### 2021-05-12 paper/agree.tex
Add formalism to AGREE semantics

@@

The revised AGREE specification for the SW component with its primary inputs and outputs is in \figref{fig:sw}. Without going into too much detail, the language is a first order predicate calculus. The semantics are synchronous data-flow where the inputs, outputs, and expressions are characterized by data streams. The semantics are such that the contracts are evaluated in dependency order with inputs being propagated to outputs through all the contracts until they stabilize; as such, the contracts, and thereby the system model, must be acyclic. Once the contracts have stabilized, then the model takes a synchronous step to the next input data in the stream. The semantics do not model computation or communication delay. The output of one contract is seen at the input of any downstream contract in the same step of the input data stream.

The contract in \figref{fig:sw} uses \texttt{eq} statements to define variables local to the contract. For example, the \texttt{req} variable is equivalent to the \texttt{event(AutomationRequest)} expression. There is an implicit \emph{event} input (or output) associated with every named event port in a component. The semantics used here do not not buffer these events so the implicit input (or output) is only a boolean value. An \texttt{event} expression refers to that implicit input (or output) and is true if there is data on the named port. The system contract assumes well-formed input and guarantees a set of properties about the output.

The guarantee is an invariant on the expression \texttt{policy or since} meaning that either the policy holds or the alert is sounding. The \texttt{policy} is defined by two local values: \texttt{current} and \texttt{previous}. The \texttt{current} value is asserted when in the current time step there is a request with a response, or there is no request and no response.  The value of \texttt{previous} in the current time step relies on values from the previous time step \texttt{->} operator for initialization as the previous time step is undefined in the first step of the system. The left operand is the initial value of \texttt{previous} at start, which in this example is \texttt{(req and not rsp)}, because seeing a request with no response is inconclusive in the first step of the system. The right operand is the value of \texttt{previous} after the initial step. Here the \texttt{pre} operator refers to the value of the expression \texttt{(req and not rsp)} in the prior time step. Intuitively, \texttt{previous} is true if the previous time step made a request without a matching response and the current time step has the matching response to that request with no new request.

The value of \texttt{since} in \emph{Alert if start is not bounded relative to a request} relies on its own value in the previous time step. The intuitive reading of the expression is that the alert has been true since the time when it first sounded. The first \texttt{alrt} sets \texttt{since} to true, and once the value of \texttt{since} is true, that value persists as long as \texttt{alrt} holds. The \emph{Alert if start is not bounded relative to a request} guarantee defines one behavior of a correctly implemented system. Together with the other guarantees, the contract models the expected input and output of the system as a whole.

@@

The hardened system is implemented by inserting the high-assurance components. The AGREE specifications for both are in \figref{fig:hardened}. Both are stylized for synthesis. The stylized filter form as generated by the tool is shown in \figref{fig:assurance}(a). The user must declare the meaning of the \texttt{policy} value as part of the transform on the model. In this example, \texttt{policy} is the assumption made by the waypoint manager about the automation response being well-formed. The output guarantees are automatically generated by the tool. The monitor is shown in \figref{fig:assurance}(b). The \texttt{is\_latched} value makes the alert persistent meaning that once raised always raised. The definition for \texttt{policy} is taken from the contract in \figref{fig:sw}. As before the guarantees for the outputs are autogenerated by the tool.

Add text to this section that defines the meaning of the contract over streams. State the prove guarantee from AGREE is terms of the streams. The statement will be used in the synthesis section to argue correctness.

>>>>

The goal of this section is to illustrate in more detail the process a system designer follows to add cyber requirements to the AGREE specifications and then automatically transform the system to insert the high-assurance components. The AGREE specifications generated by the transforms for the high-assurance components are explained. The section ends with a concise formal statement of the meaning of an AGREE specification for a high-assurance component. This meaning is what must be preserved by synthesis.

@@

The AGREE specification for the SW component in the example with the added cyber requirements is given in \figref{fig:sw}. Without going into too much detail, the language is a first order predicate calculus. The semantics are synchronous data-flow where the inputs, outputs, and expressions are characterized by data streams that comply with the input assumptions. The semantics are such that the contracts are evaluated in dependency order with inputs being propagated to outputs through all the contracts until they stabilize; as such, the contracts, and thereby the top-level model, must be acyclic. Once the contracts have stabilized, then the model takes a synchronous step to the next input data in the stream. The semantics do not model computation or communication delay. The output of one contract is seen at the input of any downstream contract in the same step of the input data stream.

The AGREE model checker attempts to prove several properties of the top-level model being verified. The first is that the output guarantees of each component implementing the system are strong enough to guarantee the input assumptions of any downstream component and strong enough to satisfy the guarantees of the top-level component's output that is being verified (i.e., the system composition meets input assumptions at each input and the guarantees on the system output). These properties are reported in the expanded lists in \figref{fig:example}(b) and \figref{fig:hardened}(b).  The next set of properties prove that the contract specifications for each component are self-consistent (i.e, a contract does not contradict itself). These are the unexpanded results to the same two figures.

Returning back to the contract in \figref{fig:sw}, it uses \texttt{eq} statements to define variables local to the contract specification. For example, the \texttt{req} variable is equivalent to the \texttt{event(AutomationRequest)} expression. In the AGREE semantics, there is an implicit \emph{event} input (or output) associated with every named event port in a component. The semantics used here do not not buffer these events so the implicit input (or output) is only a boolean value. An \texttt{event} expression refers to that implicit input (or output) and is true if there is data on the named port. The system contract here states assumptions on well-formed input, and it then states guarantees on properties about the output.

The \emph{Alert if start is not bounded relative to a request} guarantee is an invariant on the expression \texttt{policy or since} meaning that either the policy holds or the alert is sounding. The \texttt{policy} is defined by two local values: \texttt{current} and \texttt{previous}. The \texttt{current} value is asserted when in the current time step there is a request with a response, or there is no request and no response.

The value of \texttt{previous} in the current time step relies on values from the previous time step. The \texttt{->} operator is for initialization as the previous time step is undefined in the first step of the system. The left operand to the operator is the initial value of \texttt{previous} at start, which in this example is \texttt{(req and not rsp)}, because seeing a request with no response is inconclusive in the first step of the system. The right operand is the value of \texttt{previous} after the initial step. Here the \texttt{pre} operator refers to the value of the expression \texttt{(req and not rsp)} in the prior time step. Intuitively, \texttt{previous} is true if the previous time step made a request without a matching response and the current time step has the matching response to that request with no new request.

The value of \texttt{since} in \emph{Alert if start is not bounded relative to a request} relies on its own value in the previous time step. The intuitive reading of the expression is that the alert has been true since the time when it first sounded. The first \texttt{alrt} sets \texttt{since} to true, and once the value of \texttt{since} is true, that value persists as long as \texttt{alrt} holds. The \emph{Alert if start is not bounded relative to a request} guarantee defines one requirement of a cyber-hardened system implementation. Together with the other guarantees, the contract models the expected input and output of the system as a whole.

\begin{figure}
  \begin{center}
    \begin{tabular}{c}
      \includegraphics[scale=0.3]{dialogue.png}
    \end{tabular}
  \end{center}
  \caption{Dialogue for automatically transforming the model with a filter.}
  \label{fig:dialogue}
\end{figure}

@@

As noted previously, the original system fails to guarantee the cyber requirement. BriefCASE provides two transformations to address the failing requirements: inserting a filter and inserting a monitor. The component is added by selecting the connection in the model where the high-assurance component is to be added, and then choosing the appropriate transformation. The dialogue or a filter transformation is shown in \figref{fig:dialogue}. The system designer can state the policy directly in the dialogue, or leave it  blank. In this example, the policy is specified as \texttt{WELL\_FORMED\_AUTOMATION\_RESPONSE(Input)}. Additionally the filter can be attached to any requirement the vulnerability analysis. This association is useful the the assurance case from the Resolute tool [CITE DESTION 2021]. The \emph{OK} button creates the component and inserts it accordingly into the implementation

The AGREE contract specification generated by the transform is shown in \figref{fig:assurance}(a). The guarantee is stylized for synthesis and completely defines the meaning of the output under all possible input. As similar dialogue exist for inserting monitors that allows the system designer to add and remove inputs as needed. The resulting AGREE specification for the monitor in this example is shown in \figref{fig:assurance}(b). The \texttt{is\_latched} value makes the alert persistent meaning that once raised always raised and is one of the several options available in the dialogue. The definition for \texttt{policy} is taken by the system designer from the contract in \figref{fig:sw}. As before the guarantees for the outputs are autogenerated by the tool and completely define each output under every possible input.

At this point it is important to state the meaning of a high-assurance component more formally. Assume that the inputs to an AGREE contract specification are raw strings (e.g., contiguous sequences of bytes) for each named port of the component such as what might be sent over a wire by the communication fabric. An \emph{environment}, $\theta: \lval \mapsto \konst{string}$, divides the raw string into its constituent parts as given by its data definition in AADL. As such, it binds \emph{L-values} to strings where an L-value is a concept from compilers and represents anything that can appear on the left-hand side of an assignment. In the context of this presentation, an $\lval$ is any reference to a named port, field in a record, entry in an array, local value defined by an eq-statement etc. The environment maps any L-value in the scope of the AGREE contract specification to its associated raw representation as a subset of bytes in a contiguous string.

Let $s$ be the AGREE contract specification for a high-assurance component. The function, $\konst{eval}\ s\ \theta\dots$, evaluates $s$ on the given stream of total environments, $\theta\dots$, and returns \konst{true} if the specification is invariant along the entire stream and \konst{false} otherwise. The meaning of invariance for the specification merits some discussion. A guarantee has its intuitive meaning: the truth value of the associated expression at every step of the stream must be preserved. An eq-statement, however, is treated as a guarantee rather than a computation that adds a new binding to the environment. It asserts an equivalence between the named value and its associated expression. That equivalence must be preserved at every step of the stream.

The set $\Theta_s$ is the universe of environment bindings for the given specification $s$. 
The meaning of an AGREE contract specification is now defined as follows.
%
\[
  \mathcal{L}(s) = \{\theta\ldots \in \Theta_s^* \mid \konst{eval}\ s\ \theta\ldots\}  
\]
%
The meaning is the prefix closed set of finite environment streams on which the specification is invariant. Each environment in the stream is total meaning that it binds all L-values in the scope if the specification.

This paper claims that a synthesized high-assurance component generates $\mathcal{L}(s)$ precisely. Let $\konst{synth}(s)$ be the CakeML synthesized component, and let $\theta^\prime = \konst{eval}\ \konst{synth}(s)\ \theta\ldots$ be a function that evaluates the synthesized component on the input stream to produce a corresponding output stream that is total in its bindings. Let $\Theta_I$ be the universe of environments that only bind the input for the specification. This paper claims the following.
%
\[
  \forall \theta \in \Theta_I^*\ (\konst{eval}\ \konst{synth}(s)\ \theta\ldots) \in \mathcal{L}(s)) 
\]
%
As noted previously, the semantics of the AGREE specification language are understood in terms of streams of data.
